SunPy - Python for Solar Physics

### 2014-01-30 1-Introduction.tex

Science is driven by the analysis of data of growing variety and complexity.
Modern advances in sensor
technology, combined with the availability of inexpensive storage, has led to
rapid increases in the amount of data available to scientists in almost
every discipline.  Solar physics is no exception to this trend. For example,
NASA's \textit{Solar Dynamics Observatory} (\textit{SDO}) spacecraft, launched 
in February 2010, produces over 1 TB of data per day \citep{aia}. Managing and
analysing this mountain of data requires increasingly sophisticated software 
tools.
These tools should be robust, easy of use and modify, have a transparent 
development history, as well as conforming to modern software engineering 
standards.
Software with these qualities provide a strong foundation which can support the 
needs of the community as data volumes grow and science questions evolve.

The SunPy project aims to provide a free, open-source and openly developed 
software package for the analysis and visualisation of solar data. SunPy makes 
use of the Python programming language and scientific Python packages. Python 
is a general-purpose, powerful and easy-to-learn high-level programming 
language.
% RJH: Can we cut the next sentence?  It seems superfluous and top 10 is rather
% meaningless.
% schriste - i think it's important. it contrasts nicely with IDL and its tiny 
%community.
%SJM: I think it should stay, it makes a point.
%According to the 2014 TIOBE 
%Index\footnote{\url{http://www.tiobe.com/index.php/content/paperinfo/tpci/index.html}},
%it is one of the top ten most popular programming languages in the world 
%removed until better reference is found for this statistic.
Python is widely used outside of scientific fields in areas like web development, 
education and `big data' analytics.

The development of a package such as SunPy is made possible by the 
rich ecosystem of scientific packages available in Python. Core packages such 
as \texttt{NumPy}, \texttt{SciPy} and \texttt{matplotlib} 
provide the basic functionality expected of a scientific programming language:
e.g., array manipulation, core numerical algorithms and visualisation. 
Building upon these foundations, packages such as \texttt{astropy}, \texttt{pandas} and 
\texttt{scikit-image} provide more domain-specific functionality.

SunPy is designed to be a clean, simple-to-use, and well-structured 
package that provides the \textit{core} tools for solar data analysis. 
The primary focus of SunPy's early development are 
specialised data types that enable the acquisition, processing and 
visualisation of all types of solar data.

>>>>

Science is driven by the analysis of data of growing variety and complexity.
Modern advances in sensor technology, combined with the availability of inexpensive 
storage, has led to rapid increases in the amount of data available to scientists in almost
every discipline.  Solar physics is no exception to this trend. For example,
NASA's \textit{Solar Dynamics Observatory} (\textit{SDO}) spacecraft, launched
in February 2010, produces over 1 TB of data per day \citep{aia}. Managing and
analysing this mountain of data requires increasingly sophisticated software
tools. These tools should be robust, easy to use and modify, have a transparent
development history, as well as conforming to modern software engineering
standards. Software with these qualities provide a strong foundation which can support the
needs of the community as data volumes grow and science questions evolve.

The SunPy project aims to provide a free, open-source and openly developed
software package for the analysis and visualisation of solar data. SunPy makes
use of Python and scientific Python packages. Python is a free, general-purpose, 
powerful and easy-to-learn high-level programming language. Additionally, Python is 
widely used outside of scientific fields in areas like `big data' analytics, web 
development and in educational environments. For example, \texttt{pandas} was 
originally developed for quantitative analysis of financial data and has since 
grown into a generalised time-series data-analysis package. Python continues to 
see increased use in the astronomy community \citep{greenfield2011}, which has 
similar goals and requirements as the solar physics community. Finally, Python 
integrates well with many technologies such as web servers \citep{dolgert2008} and databases.

The development of a package such as SunPy is made possible by the rich ecosystem of 
scientific packages available in Python. Core packages such as \texttt{NumPy}, 
\texttt{SciPy} and \texttt{matplotlib} provide the basic functionality expected of a 
scientific programming language,
such as array manipulation, core numerical algorithms, and visualisation.
Building upon these foundations, packages such as \texttt{astropy}, \texttt{pandas} and
\texttt{scikit-image} provide more domain-specific functionality.

SunPy is designed to be a clean, simple-to-use, and well-structured package that provides 
the \textit{core} tools for solar data analysis, motivated by the need for a free and 
modern alternative to the existing SolarSoft (SSW) library. While SSW is open source 
and freely available, it relies on IDL (Interactive Data Language), a proprietary 
data-analysis environment.

### 2014-01-30 3-3-HELIO.tex

SunPy's \texttt{hec} module provides an interface to the
\textit{HELIO Event Catalogue} (HEC) service. 
%schriste - why is this italics?
This module was developed as
part of a Google Summer of Code (GSOC) project in 2013.
The HEC service currently provides access to 84 catalogues from different
sources.
As with all of the HELIO services, the HEC service provides results in VOTable 
data format (defined by IVOA \citep{ochsenbein_ivoa_2011}).
The \texttt{hec} module parses this output using the \texttt{astropy.io.votable} package.
This format has the advantage of containing metadata with information like
data provenance and the performed query.

@@

In the future, additional HELIO services will be supported by SunPy, aiming for
seamless interaction between the other services and tools available (e.g., 
\texttt{hek}, \texttt{map}).  
% RJH: This sentence is an orphan and should have a home constructed for it.
% schriste - it also contains too much future sauce
% better? or this should be in the future section?
% JI - moved bit about HELIO overplotting to Future.tex.  Might want to have
% a bit in the future section about other net services.

>>>>

SunPy's \texttt{hec} module provides an interface to the
HELIO Event Catalogue (HEC) service. 
This module was developed as
part of a Google Summer of Code (GSOC) project in 2013.
The HEC service currently provides access to 84 catalogues from different
sources.
As with all of the HELIO services, the HEC service provides results in VOTable 
data format (defined by IVOA \citep{ochsenbein_ivoa_2011}).
The \texttt{hec} module parses this output using the \texttt{astropy.io.votable} package.
This format has the advantage of containing metadata with information like
data provenance and the performed query.

@@

### 2014-01-30 4-Util.tex

%Two modules are described in this Section, namely, solar coordinates (WCS), and 
%the Sun module.

@@

%ays: I removed this as a separate subsection because constants is part of sun...
%\subsubsection{constants}\label{ssec:util:const}
The \texttt{sun.constants} module provides a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a \texttt{Constant} object as defined by Astropy. Each 
\texttt{Constant} object defines a \texttt{Quantity}, a number associated with a unit, along with 
the constant's provenance (i.e., reference) as well as its uncertainty. Using 
Astropy's {Quantity} objects, any solar constant can easily be converted between 
different units as can be seen in Listing~\ref{code:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
NumPy's \texttt{ndarray}, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the module. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly, a \texttt{print\_all()} function is available.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.sun import constants
>>> print(sun.mass)
  Name   = Solar mass
  Value  = 1.9891e+30
  Error  = 5e+25
  Units  = kg
  Reference = Allen's Astrophysical Quantities 4th Ed.
# Verify the average density of the Sun and convert to cgs
>>> (constants.mass/constants.volume).cgs
<Quantity 1.40851154227 g / (cm3)>
# Search for the age of the Sun
>>> constants.find('age')
['age', 'average angular size', 'average density', 'average intensity']
>>> constants.value('age'), constants.unit('age')
(4600000000.0, Unit("yr"))
\end{minted}
\caption{Using the \texttt{sun.constants} module.}
\label{code:constants_code}
\end{listing}

>>>>

@@

The \texttt{sun.constants} module provides a number of solar-related 
constants in order to provide consistency in the calculations of derived solar 
values both within the SunPy code base but also to the user. Every solar 
constant is provided as a \texttt{Constant} object as defined by Astropy. Each 
\texttt{Constant} object defines a \texttt{Quantity}, a number associated with a unit, along with 
the constant's provenance (i.e., reference) as well as its uncertainty. Using 
Astropy's {Quantity} objects, any solar constant can easily be converted between 
different units as can be seen in Listing~\ref{code:constants_code} as well
as between the SI or cgs unit systems with a single command.  As these objects inherit from 
NumPy's \texttt{ndarray}, they work well with standard representations of numbers.
For convenience a number of shortcuts to frequently used constants are provided 
directly when importing the module. A larger list of constants can be 
accessed through an interface modelled on that provided by the SciPy constants 
package and is available as a dictionary called \texttt{physical\_constants}. 
To view them all quickly, a \texttt{print\_all()} function is available.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.sun import constants
>>> print(constants.mass)
  Name   = Solar mass
  Value  = 1.9891e+30
  Error  = 5e+25
  Units  = kg
  Reference = Allen's Astrophysical Quantities 4th Ed.
# Verify the average density of the Sun and convert to cgs
>>> (constants.mass/constants.volume).cgs
<Quantity 1.40851154227 g / (cm3)>
# Search for the age of the Sun
>>> constants.find('age')
['age', 'average angular size', 'average density', 'average intensity']
>>> constants.value('age'), constants.unit('age')
(4600000000.0, Unit("yr"))
\end{minted}
\caption{Using the \texttt{sun.constants} module.}
\label{code:constants_code}
\end{listing}

### 2014-01-30 5-Dev.tex

\section{Development and Community}\label{sec:dev}
SunPy is a community-developed library, designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence, the whole 
development process is also online and open for anyone to contribute to.
SunPy's development makes use of the online service 
\href{http://github.com}{GitHub} and \href{http://git-scm.com/}{Git}
as its distributed version control software.

The continued success of an open-source project depends on many factors;
three of the most important are (1) utility and quality of the code, (2) documentation, and (3) an
active community \citep{bangerth2013}. Several tools (some specific to Python) are used by
SunPy to make achieving these goals more accessible. To maintain high-quality code, a 
transparent and collaborative development workflow made possible by GitHub is used.
% JI - I don't think we need a description of what is essentially how a project works
% when you use git and github.  There is nothing SunPy specific about this.
The following conditions typically must be met before code is accepted.
\begin{enumerate}
	\item  The code must follow the
	\href{http://www.python.org/dev/peps/pep-0008/}{PEP 8} Python style 
	guidelines to maintain consistency in the SunPy code.

@@

	\item The code must contain significant unit-test coverage.

@@

Additionally, SunPy makes use of `continuous integration' provided by
\href{http://travis-ci.org}{Travis CI}, a process by which the addition of any new code 
automatically triggers a comprehensive review of written unit tests. If any single test
fails, the community is alerted before the code is accepted. The unit test coverage is
also monitored by the \href{http://coveralls.io}{Coveralls} service.
% JI - what is the coveralls service?

High-quality documentation is
one of the most important factors determining the success of any software project. 
Powerful tools already exist in Python to support documentation, thanks to native
Python's focus on its own documentation. SunPy makes use of the \href{http://sphinx-doc.org}{Sphinx}
documentation generator. Sphinx uses reStructuredText as its markup language, which is
an easy-to-read, what-you-see-is-what-you-get plaintext markup syntax. It supports
many output formats most notably HTML, as well as PDF and ePub, and provides a rich,
hierarchically structured view of in-code documentation strings. The SunPy documentation 
is built automatically and is hosted by \href{http://readthedocs.org}{Read the 
Docs} at \url{http://docs.sunpy.org}.

Communication is the key to maintaining an active community, and the SunPy community 
uses a number of different tools to facilitate communication. For immediate communications, an active IRC chat
room (\#SunPy) is hosted on freenode.net. For more involved or less immediate needs, such as
developer comments or discussions, a open mailing list is hosted on Google Groups. 
Bug tracking, code reviews, and feature-request discussions take place directly on GitHub.
The SunPy community also reaches out to the wider solar physics
community through presentations, functionality demonstrations and informal meetups at scientific
meetings. 
% RJH: do we want to list both the community and dev lists?

% RJH: This section is redundant with some parts in the text.  I favor cutting the stuff in text.
% schriste - we had agreed to only refer to the support in one place. we need 
%to agree as to where
% SJM: Drop the bits in the code
Many of the larger features in SunPy have been developed 
with the support of both the ESA-SOCIS and GSOC. 
Initial development of SunPy's VSO and HEK 
implementations were funded by the SOCIS program as well as a prototyped GUI 
and N-Dimensional datatype implementations. SunPy was supported by GSOC in 2013 
through the \href{http://www.python.org/psf/}{Python Software Foundation} (PSF). It supported development of the 
\texttt{helio} and 
\texttt{database} subpackages.

%JI The section ``Development workflow'' is unnecessary IMHO.  This is
% largely a description of how git and github works.  It is not about sunpy. So I have
% deleted it.

>>>>

\section{Development and Community}\label{sec:dev}
SunPy is a community-developed library, designed and developed for and by 
the solar physics community. Not only is all the source code publicly available 
online under the permissive 2-clause BSD licence, the whole 
development process is also online and open for anyone to contribute to.
SunPy's development makes use of the online service 
GitHub (\url{http://github.com}) and Git\footnote{For more information see \url{http://git-scm.com/}}
as its distributed version control software.

The continued success of an open-source project depends on many factors;
three of the most important are (1) utility and quality of the code, (2) documentation, and (3) an
active community \citep{bangerth2013}. Several tools (some specific to Python) are used by
SunPy to make achieving these goals more accessible. To maintain high-quality code, a 
transparent and collaborative development workflow made possible by GitHub is used.
The following conditions typically must be met before code is accepted.
\begin{enumerate}
	\item  The code must follow the
	\href{http://www.python.org/dev/peps/pep-0008/}{PEP 8} Python style 
	guidelines to maintain consistency in the SunPy code.

@@

	\item The code must contain unit tests to verify that the code is behaving 
	as expected.

@@

Additionally, SunPy makes use of `continuous integration' provided by Travis CI (\url{http://travis-ci.org}), a process by which the addition of any new code 
automatically triggers a comprehensive review of the code functionality which are maintained as unit tests.
 If any single test
fails, the community is alerted before the code is accepted. The unit-test coverage is monitored by
a service called Coveralls (\url{http://coveralls.io}).

High-quality documentation is
one of the most important factors determining the success of any software project. 
Powerful tools already exist in Python to support documentation, thanks to native
Python's focus on its own documentation. SunPy makes use of the Sphinx (\url{http://sphinx-doc.org})
documentation generator. Sphinx uses reStructuredText as its markup language, which is
an easy-to-read, what-you-see-is-what-you-get plaintext markup syntax. It supports
many output formats most notably HTML, as well as PDF and ePub, and provides a rich,
hierarchically structured view of in-code documentation strings. The SunPy documentation 
is built automatically and is hosted by Read-the 
-Docs (\url{http://readthedocs.org}) at \url{http://docs.sunpy.org}.

Communication is the key to maintaining an active community, and the SunPy community 
uses a number of different tools to facilitate communication. For immediate communications, an active IRC chat
room (\#SunPy) is hosted on freenode.net. For more involved or less immediate needs, such as
developer comments or discussions, an open mailing list is hosted by Google Groups. 
Bug tracking, code reviews, and feature-request discussions take place directly on GitHub.
The SunPy community also reaches out to the wider solar physics
community through presentations, functionality demonstrations and informal meetups at scientific
meetings.
